---
title: "Testing the Stroop Effect"
author: "Mohamed Khodeir"
date: "December 12, 2015"
output: html_document
---
```{r, echo = FALSE}
library(ggplot2)
data <- read.csv('stroopdata.csv')
data$Difference <- data$Incongruent - data$Congruent
```
## References

* http://www.biostathandbook.com/pairedttest.html

## Questions

**What is our independent variable? What is our dependent variable?**

Our dependent variable is whether or not the color words match the color in which they are printed.

Our independent variable is the time taken to name the colors in equally-sized lists.

**What is an appropriate set of hypotheses for this task? What kind of statistical test do you expect to perform? Justify your choices.**

An appropriate null hypothesis is that there is no difference in the time taken to name all the colors in the list.

A good alternative hypothesis is that it takes longer to name the colors in the incongruous condition. In this case I would perform a one-tailed hypothesis test. Depending on the distribution of the data, I would use either a t-test or a non-parametric test such as the Wilcoxon signed-rank test.


**Report some descriptive statistics regarding this dataset. Include at least one measure of central tendency and at least one measure of variability.**

```{r, echo = FALSE}
writeLines(sprintf("Mean:\t\t\t\t%.3f\nMedian:\t\t\t\t%.3f\nStandard Deviation:\t%.3f",mean(data$Difference),median(data$Difference),sd(data$Difference)))
```

**Provide one or two visualizations that show the distribution of the sample data. Write one or two sentences noting what you observe about the plot or plots.***

```{r, echo = FALSE}
ggplot(data=data, aes(x=Difference)) + geom_histogram(binwidth=1.3)  + ggtitle('Histogram of Difference')
```

The limited size of the dataset makes analysing the distribution slightly problematic. However, the histogram shows some positive skew, with the right tail being longer. Also the distribution seems slightly bi-modal with two peaks.  

```{r, echo = FALSE}
qplot(sample=scale(data$Difference), stat='qq_norm') + geom_abline() + ggtitle('QQ Plot of Difference')
```

The qq-plot shows some deviations from normality. Again, we see the longer right tail and shorter left tail. 

**Now, perform the statistical test and report your results. What is your confidence level and your critical statistic value? Do you reject the null hypothesis or fail to reject it? Come to a conclusion in terms of the experiment task. Did the results match up with your expectations?**

I think it would be safer to use the non-parametric test since the assumption of normality for the difference of the paired samples does not seem reasonable. 

I will use an alpha level of .05.


```{r, echo = FALSE}
# t.test(data$Difference, alternative = 'greater')
wilcox.test(data$Difference, alternative = 'greater')
```

Since $p = 5.96e-08 < .05$ then I would have to reject the null hypothesis in favor if the alternative hypothesis. Namely, that the incongruent condition really takes the subjects more time to complete.

This result is as expected. My own time for the incongruent condition was at least 8 seconds longer. The effect of the distraction of conflicting information from text and color seems very noticable.
